---
title: "Basic R: Matrices"
author: "Nanfang Hong (U92430309)"
date: "January 27, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Matrix problems

### 1. Suppose

$$ A = \begin{bmatrix}
1 & 1 & 3\\ 
5 & 2 & 6\\ 
-2 & -1 & -3
\end{bmatrix}$$
(a) Check that $A^3 = \mathbf{0}$  
```{r}
A <- matrix(c(1,1,3,5,2,6,-2,-1,-3), nrow = 3, byrow = TRUE)
A
A %*% A %*% A
```

(b) Replace the third column of A by the 
sum of the second and third columns  

  

```{r}
A[,3] <- A[,2] + A[,3]
A
```



### 2. Create the following matrix B with 15 rows

$$B = \begin{bmatrix}
10 & -10 & 10\\
10 &  -10& 10\\
 \cdots & \cdots  & \cdots \\
10 & -10 & 10
\end{bmatrix} $$

Calculate the 3x3 matrix $B^TB$.  You can make this calculation with the function crossprod().  See the documentaion. 

```{r}
B <- matrix(c(10, -10, 10), 15, 3, byrow = TRUE)
B
crossprod(B)
```


### 3. Create a 6 x 6 matrix matE  with every element equal to 0.  check what the functions row() and col() return when applied to matE.


$$\begin{bmatrix}
0 &1  & 0 & 0 & 0 & 0\\
1 &0 & 1 &0  & 0 &0 \\
0 & 1 & 0 & 1 & 0 & 0\\
0 & 0 & 1 & 0 & 1 & 0\\
0& 0 & 0 & 1 & 0 & 1\\
0& 0 & 0 & 0 & 1 & 0
\end{bmatrix} $$  


```{r}
matE <- matrix(rep(0,36), nrow = 6, byrow = TRUE)  

```


```{r}
# With a little experimentation you would see
# that the specified pattern is in the |1|'s
row(matE)-col(matE)
```

```{r}
# so you use the locations of the 1's to modify matE 
matE[abs(row(matE)-col(matE))==1] <- 1
matE
```



### 4.  Look at the help for the function outer().  Now, create the following patterned matrix:

$$\begin{bmatrix}
0 & 1 & 2 & 3 & 4\\
1 & 2 & 3 & 4 & 5\\
2 & 3 & 4 & 5 & 6\\
3 & 4 & 5 & 6 & 7\\
4 & 5 & 6 & 7 & 8
\end{bmatrix} $$

```{r}

a <- 0:4
A <- outer(a,a,"+")
A
```



### 5.  Create the following patterned matrices.  Your solutions should be generalizable to enable creating larger matrices with the same structure.

(a)

$$\begin{bmatrix}
0 & 1 & 2 & 3 & 4\\
1 &2 & 3 & 4 & 0\\
2 & 3 & 4 & 0 & 1\\
3 & 4 & 0 & 1 & 2\\
4 & 0 & 1 & 2 & 3
\end{bmatrix}  $$
```{r}
a <- 0:4
A <- outer(a,a,"+")
A <- A %% 5
A
```

(b)

$$ \begin{bmatrix}
0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9\\
1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 0\\
\vdots  & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
8 & 9 & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7\\
9 & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8
\end{bmatrix} $$
```{r}
a <- 0:9
A <- outer(a,a,"+")
A <- A %% 10
A
```


(c)

$$\begin{bmatrix}
0 & 8 & 7 & 6 & 5 & 4 &3  & 2 & 1\\
1 & 0 & 8 & 7 & 6 & 5 & 4 & 3 & 2\\
2 & 1 & 0 & 8 & 7 & 6 & 5 & 4 & 3\\
3 & 2 & 1 & 0 & 8 & 7 & 6 & 5 & 4\\
4 & 3 & 2 & 1 & 0 & 8 & 7 & 6 & 5\\
5 & 4 & 3 & 2 & 1 & 0 & 8 & 7 & 6\\
6 & 5 & 4 & 3 & 2 & 1 & 0 & 8 & 7\\
7 & 6 & 5 & 4 & 3 & 2 & 1 & 0 & 8\\
8 & 7 & 6 & 5 & 4 & 3 & 2 & 1 & 0
\end{bmatrix} $$
```{r}
a <- 0:8
b <- 9:1
A <- outer(a,b,"+")
A <- A %% 9
A
```


### 6.  Solve the following system of linear equations by setting up and solving the matrix equation Ax = y.

$x_1 + 2x_2 + 3x_3 + 4x_4 + 5x_5 =  7$  
$2x_1 +  x_2 + 2x_3 + 3x_4 + 4x_5 = -1$  
$3x_1 + 2x_2 +  x_3 + 2x_4 + 3x_5 = -3$  
$4x_1 + 3x_2 + 2x_3 +  x_4 + 2x_5 =  5$  
$5x_1 + 4x_2 + 3x_3 + 2x_4 +  x_5 = 17$  

```{r}
A <- matrix(rep(0, 25), nrow = 5, byrow = TRUE)  
A <- abs(row(A)-col(A)) + 1
y <- c(7, -1, -3, 5, 17)
solve(A, y)
```

### 7. Create a 6 x 10 matrix of random integers chosen from 1,2,...,10 by executing the following two lines of code:

```{r}
set.seed(75)  
aMat <- matrix(sample(10, size=60, replace=TRUE), nr=6)  
aMat
```


Use the matrix you have created to answer these questions:  

(a) Find the number of entries in each row which are greater than 4.

```{r}
rowSums(ifelse(aMat > 4, TRUE, FALSE))
```



(b) Which rows contain exactly two occurrences of the number seven?

```{r}
rowsum1 <- rowSums(ifelse(aMat == 7, TRUE, FALSE))
names(rowsum1) <- paste("Row", 1:length(rowsum1))
print(names(rowsum1[rowsum1 == 2]))
```


(c) Find those pairs of columns whose total (over both columns) is greater tnan 75.  The answer should be a matrix with two columns; so, for example, the row (1,2) in the output matrix means that the sum of columns 1 and 2 in the original matrix is greater than 75.  Repeating a column is permitted; so, for exalmple, the final output matrix could contain the rows (1,2), (2,1), and (2,2).  

```{r}
# Ask Dr. Brian what the output is desired?
colsum1 <- colSums(aMat)
outPutMat <- outer(colsum1, colsum1, "+")
outPutMat
index <- which(outPutMat > 75)
colMat <- matrix(c(ceiling(index / 10), index %% 10), ncol = 2)
colMat
```


What if repetitions are not permitted? Then only (1,2) from (1,2),(2,1) and (2,2) would be permitted.
```{r}
del <- which((colMat[ ,1] < colMat[ ,2]) == FALSE)
colMat[-del, ]
```


### 8. Calculate

(a) $\sum\limits_{i=1}^{20} \sum\limits_{j=1}^5 \frac{i^4}{(3+j)}$


```{r}
sum(sapply(1:20, function(i) (i ^ 4) / (3 + (1:5))))

```


(b) $\sum\limits_{i=1}^{20} \sum\limits_{j=1}^5 \frac{i^4}{(3+ij)}$

```{r}
sum(sapply(1:20, function(i) (i ^ 4) / (3 + i * (1:5))))
```

(c) $\sum\limits_{i=1}^{10} \sum\limits_{j=1}^i \frac{i^4}{(3+ij)}$

```{r}
a <- sapply(1:20, function(i) (i ^ 4) / (3 + i * (1:i)))
sum(sapply(1:20, function(x) sum(a[[x]])))
```

